### 2022-11-26 dissertation/chapters/05-probability.tex
🚧 (final-review): Chapter 5 WIP

Probability theory and statistics can be traced back to as early as the 18th century. In 1718,~\citeauthor{ref:demoivre:1718}~\cite{ref:demoivre:1718} published the \textit{Doctrine of of Chance}; a book that is widely seen as the first published book on probability theory. In 1763, Thomas Bayes~\cite{ref:bayes:1763} published an article titled \textit{An Essay towards solving a Problem in the Doctrine of Chances} where the first version of \index{Bayes' Theorem}Bayes' Theorem was introduced.

Probability theory, statistics and \acf{ML} are transdisciplinary fields with many shared concepts. There are many examples of how probability theory has been incorporated into \acs{ML} research. In 1991,~\citeauthor{ref:denker:1991},~\cite{ref:denker:1991} proposed a way to transform \acf{ANN} outputs to probability distributions. In 1993, \citeauthor{ref:neal:1993}~\cite{ref:neal:1993} developed a \acf{MCMC} sampling algorithm for Bayesian \acp{NN}. These are but a few examples of the role that probability theory has played in \acs{ML} research in the past.

Chapter \ref{chap:hhs} provided the reader with the concept of a \acf{HH}. This dissertation aims to develop a selection \acs{HH} that makes use of probability theory to select the best \acs{HH} to solve a given problem. This chapter aims to provide the reader with the necessary background information on probability theory and statistics. These are big fields and focus is put on the elements that are required to formulate the proposed \acf{BHH}. The remainder of the chapter is structured as follows:

@@

      \item \textbf{Section \ref{sec:probability:bayes_theorem}} introduces \index{Bayes' Theorem}Bayes' Theorem, the fundamental theorem upon which the \acs{BHH} is built.

@@

In everyday conversation, the term \textit{probability} is a measure of one's belief in the occurrence of a future event~\cite{ref:wackerly:2014}. Probability is a necessary tool used in many fields including physics, biology, chemistry and computer science. These fields contain many cases that generate observations that cannot be predicted with absolute certainty~\cite{ref:wackerly:2014}. Probability can be be inferred and confirmed through past events. These events are referred to as \textit{random} or \textit{stochastic} events. The probability that a certain event $A$ might occur, is denoted $P(A)$. Although these random events cannot be predicted with absolute certainty, the relative frequency with which they occur over many trials, is often remarkably stable.

Consider flipping an unbiased, fair coin. The coin has two possible outcomes. It can conclude that each side has a $\frac{1}{2}$ or $50\%$ chance of occurring. In statistics, the decimal probability notation is used, where $0 <= P(A) <= 1$. Suppose the fair coin is thrown 10 times, one is not guaranteed to observe $0.5$ heads and $0.5$ tails. There is some probability (although small) that the coin might fall on heads $0/10$ times. The probability of such an event occurring is $0.0009765625$. In the coin flip example, the \acf{CLT} shows that the normalised sum of events tends toward a normal distribution with a mean value of $0.5$ if the number of events observed $N$, is large~\cite{ref:wackerly:2014}. The larger the value of $N$, the higher the confidence of mean probability and relative frequency of the event. The stable long-term relative frequency by which a random event occurs, provides an intuitive and meaningful measure of belief that a certain event will occur again at some point in the future~\cite{ref:wackerly:2014}.

Probability can also be expressed over multiple random events. Multiple random events can be considered together, dependently or conditionally. The following sections provides insight into conditional and joints probabilities of multiple random events.

@@

The occurrence of a given random event $A$ can often be conditional on the occurrence of another event $B$. In the field of medicine, an example of this is to calculate the probability of a certain diagnosis of a sick patient given his/her symptoms. The aforementioned relationship between events is referred to as the conditional probability between two events. The conditional probability is expressed as $P(A \vert B)$ and is read \textit{the probability of $A$ given $B$}. On the contrary, the \textit{unconditional} probability is the probability of an event, not dependent on any other. The conditional probability of $A$ given $B$ can be expressed as is given in Definition \ref{eq:probability:cond_probability} below~\cite{ref:wackerly:2014}.
\\
\begin{definition}[\textbf{Conditional Probability}]
      \label{eq:probability:cond_probability}
      The conditional probability of an event $A$, subject to the occurrence of event $B$ is expressed as

@@

It could also be the case that the outcome of observing event $A$ is not affected by the occurrence of $B$. In this case, it is said that events $A$ and $B$ are independent. The independence between two events are expressed in Definition \ref{def:probability:cond_probability:independence} below.
\\
\begin{definition}[\textbf{Independence of Events}]
      \label{def:probability:cond_probability:independence}
      Two events, $A$ and $B$ are said to be independent of each other if, and only if the following criteria hold:

@@

      otherwise, events $A$ and $B$ are said to be \textit{dependent} random events.
\end{definition}

@@

\begin{proof}
      Proof is given from Definition \ref{eq:probability:cond_probability} above.
\end{proof}
\vspace*{0.5cm}

@@

\begin{proof}
      The geometric proof of the additive law of probability is given by the Venn Diagram presented in Figure \ref{fig:probability:multiple_events:additive}. Note that $A \cup B = A \cup ( \bar{A} \cap B)$, where $A$ and $\bar{A} \cap B$ are mutually exclusive events. Furthermore, consider that $B = (\bar{A} \cap B) \cup (A \cap B)$, where $\bar{A} \cap B$ and $A \cap B$ are mutually exclusive. Then $P(A \cup B) = P(A) + P(\bar{A} \cap B)$ and $P(B) = P(\bar{A} \cap B) + P(A \cap B)$. The equality on the right implies that $P(\bar{A} \cap B) = P(B) - P(A \cap B)$. By substituting the expression for $P(\bar{A} \cap B)$ into the expression for $P(A \cup B)$, one can obtain the result $P(A \cup B) = P(A) + P(B) - P(A \cap B)$.
\end{proof}

@@

\index{Bayes' Theorem}Bayes' Theorem, named after Thomas Bayes, describes the probability of an event $A$, based on prior knowledge of conditions that might be related to $A$~\cite{ref:zalta:2015}. In order to derive the formal theorem and proof, first consider Definition \ref{def:probability:bayes_theorem:partition} below~\cite{ref:zalta:2015}.
\\
\begin{definition}
      \label{def:probability:bayes_theorem:partition}
      For some positive integer, $K$, let the sets $B_{1}, B_{2}, \dots, B_{K}$ be such that

@@

\index{Bayes Theorem}
\begin{theorem}[\textbf{Bayes' Theorem}]
      \label{th:probability:bayes_theorem:theorem}
      Assume that $[B_{1}, B_{2}, \dots, B_{K}]$ is a partition of $S$ such that $P(B_{i}) > 0$, for $i = 1,2, \dots, K$ then\\
      \\
      \begin{equation}
            P(B_{j} \vert A) = \frac{P(A \vert B_{j})P(B_{j})}{\sum_{i=1}^{K} P(A \vert B_{i})P(B_{i})}
      \end{equation}
\end{theorem}

\begin{proof}
      The proof follows from the definition of conditional probability as was presented in Section~\ref{sec:probability:cond_probability}.

@@

One of the many applications of \index{Bayes' Theorem}Bayes' Theorem is to do statistical inference. \index{Bayes' Theorem}Bayes' Theorem expresses how a degree of belief, expressed as a probability, should rationally change to account for the availability of related evidence.

@@

The \index{Beta probability distribution}Beta probability distribution is a family of univariate, continuous, probability distributions over some $x$, with support on the interval $[0,1]$~\cite{ref:wackerly:2014}. It is parameterised by two shape parameters $\alpha > 0$, $\alpha \in \mathbb{R}$ and $\beta > 0$, $\beta \in \mathbb{R}$. The \index{Beta probability distribution}Beta probability distribution is denoted as $Beta(\alpha, \beta)$. The \acf{PDF} of the \index{Beta probability distribution}Beta probability distribution is given in Equation~\eqref{eq:probability:probability_distributions:beta:pdf} below.

@@

From Equation~\eqref{eq:probability:probability_distributions:beta:pdf} above, the normalising constant $B(\alpha, \beta)$, is defined below in Equation~\eqref{eq:probability:probability_distributions:beta:norm_const}.

@@

In Equation~\eqref{eq:probability:probability_distributions:beta:norm_const}, $\Gamma$ is the Gamma function as defined in Equation~\eqref{eq:probability:probability_distributions:beta:gamma_func} below.

@@

The \index{Bernoulli probability distribution}Bernoulli probability distribution is a discrete probability distribution over some random variable $x$ that takes the value of $1$ with probability $\theta$ and $0$ with probability $1-\theta$ ~\cite{ref:wackerly:2014}. The \index{Bernoulli probability distribution}Bernoulli probability distribution is denoted as $Ber(\theta)$ with support $x \in \{0, 1\}$. In probability theory, the \index{Bernoulli probability distribution}Bernoulli probability distribution is often used to explain the possible outcomes of a single experiment that asks a \textit{yes-no} question such as flipping a coin. The outcome of such an experiment is a boolean value. The \index{Bernoulli probability distribution}Bernoulli probability distribution has a \acf{PMF} as given below in Equation~\eqref{eq:probability:probability_distributions:bernoulli:pmf}.

@@

Equation~\eqref{eq:probability:probability_distributions:bernoulli:pmf} above, can also be expressed as show in Equation~\eqref{eq:probability:probability_distributions:bernoulli:alt} below.

@@

The mean of the \index{Bernoulli probability distribution}Bernoulli probability distribution approaches $\theta$ over many samples according to the \acs{CLT}~\cite{ref:grinstead:1997}. Figure \ref{fig:probability:probability_distributions:bernoulli:coin} shows a simulation of a fair-coin flipping simulation. The mean converges to 0.5 for various sample sizes.

@@

The expected value of the distribution is thus given according to Equation~\eqref{eq:probability:probability_distributions:bernoulli:expected_value} below.

@@

The mode of the distribution is given in Equation~\eqref{eq:probability:probability_distributions:bernoulli:mode} below.

\begin{equation}
      \label{eq:probability:probability_distributions:bernoulli:mode}
      M[x_{k}] =
      \begin{cases}
            0   & \text{if}\ \theta < 0.5 \\
            0,1 & \text{if}\ \theta = 0.5 \\
            1   & \text{if}\ \theta > 0.5 \\
      \end{cases}
\end{equation}

@@

The \index{Binomial probability distribution}Binomial probability distribution is a discrete probability distribution over a random variable $x$ taking on a number of successes in $N$ sequential independent experiments that each ask a \textit{yes-no} question~\cite{ref:wackerly:2014}. The probability of a single independent experiment yielding a success, is given as $\theta$ and the \index{Binomial probability distribution}Binomial probability distribution is denoted as $Bin(N, \theta)$ with support $x \in \{0, 1, \dots, N\}$.  It should be noted that the \index{Binomial probability distribution}Binomial probability distribution is the extension of the \index{Bernoulli probability distribution}Bernoulli probability distribution over $N$ independent sequential experiments and thus, each experiment also yields some boolean outcome. When $N=1$, the experiment is referred to as a Bernoulli trial, and the distribution is just a \index{Bernoulli probability distribution}Bernoulli probability distribution. When $N > 1$, the sequence of outcomes is referred to as a Bernoulli process. The \acf{PMF} of the \index{Binomial probability distribution}Binomial probability distribution is given in Equation~\eqref{eq:probability:probability_distributions:binomial:pmf} below.

@@

Similar to the \index{Bernoulli probability distribution}Bernoulli probability distribution, the mean of the \index{Binomial probability distribution}Binomial probability distribution is $N\theta$ given the \acs{CLT} as was shown in Figure \ref{fig:probability:probability_distributions:bernoulli:coin}. The expected value of the \index{Binomial probability distribution}Binomial probability distribution is thus given in Equation~\eqref{eq:probability:probability_distributions:binomial:expected_value} below.

@@

The mode of the distribution is given in Equation~\eqref{eq:probability:probability_distributions:binomial:mode} below.

\begin{align}
      \label{eq:probability:probability_distributions:binomial:mode}
      \begin{split}
            M[x_{k}] &= E[x] + \theta \\
            &= N\theta  + \theta \\
            &= (N  + 1)\theta
      \end{split}
\end{align}

@@

The \index{Categorical probability distribution}Categorical probability distribution is a discrete probability distribution over some random variable $x$, taking on any one of $K$ possible categories~\cite{ref:wackerly:2014}. There is no ordering to these categories and therefore, for simplicity, each category is assigned a numerical representative value such that $k = (1, 2, \dots, K)$. The probabilities for all outcomes is given by the probability vector $\vec{\theta} = (\theta_{1}, \dots, \theta_{K})$.  This means that the probability $P(x=k)=\theta_{k}$, has support $x \in \{1, \dots, K\}$. The \index{Categorical probability distribution}Categorical probability distribution, denoted $Cat(\theta)$, is the generalization of the \index{Bernoulli probability distribution}Bernoulli probability distribution and is sometimes referred to it by its alternative names, the generalised \index{Bernoulli probability distribution}Bernoulli probability distribution or the Multinoulli probability distribution. In probability theory, the \index{Categorical probability distribution}Categorical probability distribution is often used to explain the outcome of a single experiment with more than two possible outcomes, such as rolling a six-sided die~\cite{ref:wackerly:2014}. The \acs{PMF} of the \index{Categorical probability distribution}Categorical probability distribution is given in Equation~\eqref{eq:probability:probability_distributions:categorical:pmf} below.

@@

In Equation~\eqref{eq:probability:probability_distributions:categorical:pmf} above, $[x = k]$ is the \textit{Iversion Bracket}~\cite{ref:iverson:1962}, yielding 1 if $x = k$ and 0 otherwise. Given class $k$, the categorical distribution simply yields $\theta_{k}$ as is shown in Equation~\eqref{eq:probability:probability_distributions:categorical:pmf_k} below.

\begin{equation}
      \label{eq:probability:probability_distributions:categorical:pmf_k}
      f_{Cat}(x=k; K, \theta) = \theta_{k}
\end{equation}

The random variable $x$ can also be encoded in binary format, yielding a vector $\vec{x} = (x_{1}, \dots, x_{K})$ of \index{Bernoulli probability distribution}Bernoulli probability distributions such that the support is $\forall_{k=1}^{K} x_{k} \in \{0, 1\}$. Importantly, if the outcome of the random event is of category $k$, then $x_{k} = 1$ and $\forall_{j=1}^{K} x_{j} = 0, j \neq k$ so that the standard $K-1$ probability simplex $S$ still holds. The \acs{PMF} of the \index{Categorical probability distribution}Categorical probability distribution can then be rewritten as follows in Equation~\eqref{eq:probability:probability_distributions:categorical:pmf_alt} below.

\begin{equation}
      \label{eq:probability:probability_distributions:categorical:pmf_alt}
      f_{Cat}(x; K, \theta) = \prod_{k=1}^{K}\theta_{k}^{\mathbbm{1}_{1}(x_{k})}
\end{equation}

In Equation \ref{eq:probability:probability_distributions:categorical:pmf_alt} above, $\mathbbm{1}(x_{k})$ is the \textit{indicator function}, yielding 1 if $x_{k} = 1$ and 0 otherwise.

Since there is no order to the underlying categories, the mean of the distribution does not yield any relevant information. The mode of the distribution is given in Equation~\eqref{eq:probability:probability_distributions:categorical:mode} below.

@@

      \item Let $X = (x_{1}, \dots, x_{I})$ be the outcomes of independent, identical random events, each with boolean outcome. That is $x_{i} | \theta \overset{\text{iid}}{\sim} Ber(\theta)$ and $\mathcal{L}(x_{i} \vert \theta)$ is the Bernoulli likelihood.

@@

By \index{Bayes' Theorem}Bayes' Theorem, the posterior distribution with given prior data $\mathcal{D}$ is given in Equation~\eqref{eq:probability:conjugate_priors:binom_likelihood:posterior} below.

@@

By \index{Bayes' Theorem}Bayes' Theorem, the posterior distribution with given prior data $\mathcal{D}$ is given in Equation~\eqref{eq:probability:conjugate_priors:mult_likelihood:posterior} below.

@@

In general, there are two main views to probability and statistics. These include the \textit{frequentist} and the \textit{Bayesian} view of statistics. Naturally, Bayesian statistics is based on \index{Bayes' Theorem}Bayes' Theorem as was presented in Section \ref{sec:probability:bayes_theorem}. Bayesian statistics describe the probability of an event in terms of some belief, based on previous knowledge of the event, and the conditions under which the event happens~\cite{ref:hackenberger:2019}. To introduce the concept of Bayesian inference and analysis, one first need to understand the difference between these two approaches. Bayesian statistics out-date the frequentist approach, but lacked interest in the early days, partly because of the limited applications where the conjugate priors where known~\cite{ref:hackenberger:2019}. More recent advancements in mathematical methods popularised the Bayesian approach again. A notable contribution to this switch, was the development of \acf{MCMC} in the 1950's. This family of algorithms allowed for the construction of random sampling algorithms from a probability distribution, which allows for the calculation of Bayesian hierarchical models~\cite{ref:hackenberger:2019}. Soon after followed one of the earliest papers that use Bayesian statistics in the field of medicine in 1982~\cite{ref:ashby:2006}.

@@

Finally, the posterior density of $\theta \vert y_{1}, y_{2}, \dots, y_{n}$, according to \index{Bayes' Theorem}Bayes' Theorem, is given in Equation~\eqref{eq:probability:bayesian_statistic:bayesian_analysis:posterior_density} below.

>>>>

Probability theory and statistics can be traced back to as early as the 18th century. In 1718,~\citeauthor{ref:demoivre:1718}~\cite{ref:demoivre:1718} published the \textit{Doctrine of of Chance}; a book that is widely seen as the first published book on probability theory. In 1763, Thomas Bayes~\cite{ref:bayes:1763} published an article titled \textit{An Essay towards solving a Problem in the Doctrine of Chances} where the first version of \index{Bayes' theorem}Bayes' theorem was introduced.

Probability theory, statistics and \acf{ML} are transdisciplinary fields with many shared concepts. There are many examples of how probability theory has been incorporated into \acs{ML} research. In 1991,~\citeauthor{ref:denker:1991},~\cite{ref:denker:1991} proposed a way to transform \acf{ANN} outputs to probability distributions. In 1993, \citeauthor{ref:neal:1993}~\cite{ref:neal:1993} developed a \acf{MCMC} sampling algorithm for \acfp{BNN}. These are but a few examples of the role that probability theory has played in \acs{ML} research in the past.

Chapter \ref{chap:hhs} provided the reader with the concept of a \acf{HH}. This dissertation aims to develop a selection \acs{HH} that makes use of probability theory to select the best \acs{HH} to solve a given problem. This chapter aims to provide the reader with the necessary background information on probability theory and statistics. These are large fields and focus is put on the elements that are required to formulate the proposed \acf{BHH}. The remainder of the chapter is structured as follows:

@@

      \item \textbf{Section \ref{sec:probability:bayes_theorem}} introduces \index{Bayes' theorem}Bayes' theorem, the fundamental theorem upon which the \acs{BHH} is built.

@@

In everyday conversation, the term \textit{probability} is a measure of belief in the occurrence of a future event~\cite{ref:wackerly:2014}. Probability is a necessary tool used in many fields including physics, biology, chemistry and computer science. These fields contain many cases that generate observations that cannot be predicted with absolute certainty~\cite{ref:wackerly:2014}. Probability can be be inferred and confirmed through past events. These events are referred to as \textit{random} or \textit{stochastic} events. The probability that a certain event, $A$ might occur is denoted by $P(A)$. Although these random events cannot be predicted with absolute certainty, the relative frequency with which they occur over many trials, is often remarkably stable.

Consider flipping an unbiased, fair coin. The coin has two possible outcomes. It can conclude that each side has a $\frac{1}{2}$ or $50\%$ chance of occurring. In statistics, the decimal probability notation is used, where $0 <= P(A) <= 1$. Suppose the fair coin is thrown 10 times, there is not guarantee of observing $0.5$ heads and $0.5$ tails. There is some probability (although small) that the coin might fall on heads $0/10$ times. The probability of such an event occurring is $0.0009765625$. In the coin flip example, the \acf{CLT} shows that the normalised sum of events tends toward a normal distribution with a mean value of $0.5$ if the number of events observed $N$, is large~\cite{ref:wackerly:2014}. The larger the value of $N$, the higher the confidence of mean probability and relative frequency of the event. The stable long-term relative frequency by which a random event occurs, provides an intuitive and meaningful measure of belief that a certain event will occur again at some point in the future~\cite{ref:wackerly:2014}.

Probability can also be expressed over multiple random events. Multiple random events can be considered together, dependently or conditionally. The following sections provide insight into conditional and joint probabilities of multiple random events.

@@

The occurrence of a given random event, $A$ can often be conditional on the occurrence of another event, $B$. In the field of medicine, an example of this is to calculate the probability of a certain diagnosis of a sick patient given his/her symptoms. The aforementioned relationship between events is referred to as the conditional probability between two events. The conditional probability is expressed as $P(A \vert B)$ and is read \textit{the probability of $A$ given $B$}. On the contrary, the \textit{unconditional} probability is the probability of an event, not dependent on any other. The conditional probability of $A$ given $B$ can be expressed as is given in Definition \ref{eq:probability:cond_probability} below~\cite{ref:wackerly:2014}.
\\
\begin{definition}[\textbf{Conditional Probability}]
      \label{eq:probability:cond_probability}
      The conditional probability of an event $A$, subject to the occurrence of event $B$ is expressed as

@@

It could also be the case that the outcome of observing event $A$ is not affected by the occurrence of $B$. In this case, it is said that events $A$ and $B$ are independent. The independence between two events are expressed in Definition \ref{def:probability:cond_probability:independence} below.
\\
\begin{definition}[\textbf{Independence of Events}]
      \label{def:probability:cond_probability:independence}
      Two events, $A$ and $B$, are said to be independent of each other if, and only if the following criteria hold:

@@

      Otherwise, events $A$ and $B$ are said to be \textit{dependent} random events.
\end{definition}

@@

\begin{proof}
      Proof is given from Definition \ref{eq:probability:cond_probability}.
\end{proof}
\vspace*{0.5cm}

@@

\begin{proof}
      The geometric proof of the additive law of probability is given by the Venn Diagram presented in Figure \ref{fig:probability:multiple_events:additive}. Note that $A \cup B = A \cup ( \bar{A} \cap B)$, where $A$ and $\bar{A} \cap B$ are mutually exclusive events. Furthermore, consider that $B = (\bar{A} \cap B) \cup (A \cap B)$, where $\bar{A} \cap B$ and $A \cap B$ are mutually exclusive. Then $P(A \cup B) = P(A) + P(\bar{A} \cap B)$ and $P(B) = P(\bar{A} \cap B) + P(A \cap B)$. The equality on the right implies that $P(\bar{A} \cap B) = P(B) - P(A \cap B)$. By substituting the expression for $P(\bar{A} \cap B)$ into the expression for $P(A \cup B)$, the resulting expression $P(A \cup B) = P(A) + P(B) - P(A \cap B)$ is obtained.
\end{proof}

@@

\index{Bayes' theorem}Bayes' theorem, named after Thomas Bayes, describes the probability of an event, $A$, based on prior knowledge of conditions that might be related to $A$~\cite{ref:zalta:2015}. In order to derive the formal theorem and proof, first consider Definition \ref{def:probability:bayes_theorem:partition} below~\cite{ref:zalta:2015}.
\\
\begin{definition}
      \label{def:probability:bayes_theorem:partition}
      For some positive integer, $K$, let the sets $B_{1}, B_{2}, \dots, B_{K}$ be such that

@@

\index{Bayes Theorem}
\begin{theorem}[\textbf{Bayes' theorem}]
      \label{th:probability:bayes_theorem:theorem}
      Assume that $\{B_{1}, B_{2}, \dots, B_{K}\}$ is a partition of $S$ such that $P(B_{i}) > 0$, for $i = 1,2, \dots, K$ then\\
      \\
      \begin{equation}
            P(B_{j} \vert A) = \frac{P(A \vert B_{j})P(B_{j})}{\sum_{i=1}^{K} P(A \vert B_{i})P(B_{i})}
      \end{equation}
\end{theorem}

\begin{proof}
      The proof follows from the definition of conditional probability as was presented in Section~\ref{sec:probability:cond_probability}:

@@

One of the many applications of \index{Bayes' theorem}Bayes' theorem is to do statistical inference. \index{Bayes' theorem}Bayes' theorem expresses how a degree of belief, expressed as a probability, should rationally change to account for the availability of related evidence.

@@

The \index{Beta probability distribution}Beta probability distribution is a family of univariate, continuous, probability distributions over some $x$, with support on the interval $[0,1]$~\cite{ref:wackerly:2014}. It is parameterised by two shape parameters $\alpha > 0$, $\alpha \in \mathbb{R}$ and $\beta > 0$, $\beta \in \mathbb{R}$. The \index{Beta probability distribution}Beta probability distribution is denoted as $Beta(\alpha, \beta)$. The \acf{PDF} of the \index{Beta probability distribution}Beta probability distribution is given as

@@

The normalising constant, $B(\alpha, \beta)$, is defined as

@@

where $\Gamma$ is the Gamma function as defined in Equation~\eqref{eq:probability:probability_distributions:beta:gamma_func} below.

@@

The \index{Bernoulli probability distribution}Bernoulli probability distribution is a discrete probability distribution over some random variable $x$ that takes the value of $1$ with probability $\theta$ and $0$ with probability $1-\theta$ ~\cite{ref:wackerly:2014}. The \index{Bernoulli probability distribution}Bernoulli probability distribution is denoted as $Ber(\theta)$ with support $x \in \{0, 1\}$. In probability theory, the \index{Bernoulli probability distribution}Bernoulli probability distribution is often used to explain the possible outcomes of a single experiment that asks a \textit{yes-no} question such as flipping a coin. The outcome of such an experiment is a Boolean value. The \index{Bernoulli probability distribution}Bernoulli probability distribution has a \acf{PMF}, given as

@@

Equation~\eqref{eq:probability:probability_distributions:bernoulli:pmf} can also be expressed as

@@

The mean of the \index{Bernoulli probability distribution}Bernoulli probability distribution approaches $\theta$ over many samples according to the \acs{CLT}~\cite{ref:grinstead:1997}. Figure \ref{fig:probability:probability_distributions:bernoulli:coin} illustrates a simulation of a fair-coin flipping simulation. The mean converges to 0.5 for various sample sizes.

@@

The expected value of the distribution is thus given as

@@

The mode of the distribution is given as

\begin{equation}
      \label{eq:probability:probability_distributions:bernoulli:mode}
      M[x] =
      \begin{cases}
            0   & \text{if}\ \theta < 0.5 \\
            0,1 & \text{if}\ \theta = 0.5 \\
            1   & \text{if}\ \theta > 0.5 \\
      \end{cases}
\end{equation}

@@

The \index{Binomial probability distribution}Binomial probability distribution is a discrete probability distribution over a random variable $x$ taking on a number of successes in $N$ sequential independent experiments that each ask a \textit{yes-no} question~\cite{ref:wackerly:2014}. The probability of a single independent experiment yielding a success is given as $\theta$ and the \index{Binomial probability distribution}Binomial probability distribution is denoted as $Bin(N, \theta)$, with support $x \in \{0, 1, \dots, N\}$.  It should be noted that the \index{Binomial probability distribution}Binomial probability distribution is an extension of the \index{Bernoulli probability distribution}Bernoulli probability distribution over $N$ independent sequential experiments, and thus each experiment also yields some Boolean outcome. When $N=1$, the experiment is referred to as a Bernoulli trial, and the distribution is just a \index{Bernoulli probability distribution}Bernoulli probability distribution. When $N > 1$, the sequence of outcomes is referred to as a Bernoulli process. The \acf{PMF} of the \index{Binomial probability distribution}Binomial probability distribution is given as

@@

Similar to the \index{Bernoulli probability distribution}Bernoulli probability distribution, the mean of the \index{Binomial probability distribution}Binomial probability distribution is $N\theta$ given the \acs{CLT} as was shown in Figure \ref{fig:probability:probability_distributions:bernoulli:coin}. The expected value of the \index{Binomial probability distribution}Binomial probability distribution is thus given as

@@

The mode of the distribution is given as

\begin{align}
      \label{eq:probability:probability_distributions:binomial:mode}
      \begin{split}
            M[x] &= E[x] + \theta \\
            &= N\theta  + \theta \\
            &= (N  + 1)\theta
      \end{split}
\end{align}

@@

The \index{Categorical probability distribution}Categorical probability distribution is a discrete probability distribution over some random variable, $x$, taking on any one of $K$ possible categories~\cite{ref:wackerly:2014}. There is no ordering to these categories and therefore, for simplicity, each category is assigned a numerical representative value such that $k \in \{1, 2, \dots, K\}$. The probabilities for all outcomes is given by the probability vector $\vec{\theta} = (\theta_{1}, \dots, \theta_{K})$.  This means that the probability $P(x=k)=\theta_{k}$ has support $x \in \{1, \dots, K\}$. The \index{Categorical probability distribution}Categorical probability distribution, denoted $Cat(\theta)$, is a generalization of the \index{Bernoulli probability distribution}Bernoulli probability distribution and is sometimes referred to it by its alternative names, i.e. the generalised \index{Bernoulli probability distribution}Bernoulli probability distribution or the Multinoulli probability distribution. In probability theory, the \index{Categorical probability distribution}Categorical probability distribution is often used to explain the outcome of a single experiment with more than two possible outcomes, such as rolling a six-sided die~\cite{ref:wackerly:2014}. The \acs{PMF} of the \index{Categorical probability distribution}Categorical probability distribution is given as

@@

where the term $[x = k]$ is the \textit{Iversion Bracket}~\cite{ref:iverson:1962}, yielding 1 if $x = k$ and 0 otherwise. Given class $k$, the categorical distribution simply yields $\theta_{k}$ as is follows:

\begin{equation}
      \label{eq:probability:probability_distributions:categorical:pmf_k}
      f_{Cat}(x=k; K, \boldsymbol{\theta}) = \theta_{k}
\end{equation}

The random variable $x$ can also be encoded in binary format, yielding a vector $\boldsymbol{x} = (x_{1}, \dots, x_{K})$ of \index{Bernoulli probability distribution}Bernoulli probability distributions such that the support is $\forall_{k=1}^{K} x_{k} \in \{0, 1\}$. Importantly, if the outcome of the random event is of category $k$, then $x_{k} = 1$ and $\forall_{j=1}^{K} x_{j} = 0, j \neq k$ so that the standard $K-1$ probability simplex $S$ still holds. The \acs{PMF} of the \index{Categorical probability distribution}Categorical probability distribution is rewritten as

\begin{equation}
      \label{eq:probability:probability_distributions:categorical:pmf_alt}
      f_{Cat}(\boldsymbol{x}; K, \boldsymbol{\theta}) = \prod_{k=1}^{K}\theta_{k}^{\mathbbm{1}_{1}(x_{k})}
\end{equation}

where $\mathbbm{1}(x_{k})$ is the \textit{indicator function}, yielding 1 if $x_{k} = 1$ and 0 otherwise.

Since there is no order to the underlying categories, the mean of the distribution does not yield any relevant information. The mode of the distribution is given as

@@

      \item Let $X = (x_{1}, \dots, x_{I})$ be the outcomes of independent, identical random events, each with Boolean outcome. That is $x_{i} | \theta \overset{\text{iid}}{\sim} Ber(\theta)$ and $\mathcal{L}(x_{i} \vert \theta)$ is the Bernoulli likelihood.

@@

By \index{Bayes' theorem}Bayes' theorem, the posterior distribution with given prior data $\mathcal{D}$ is given in Equation~\eqref{eq:probability:conjugate_priors:binom_likelihood:posterior} below.

@@

By \index{Bayes' theorem}Bayes' theorem, the posterior distribution with given prior data $\mathcal{D}$ is given in Equation~\eqref{eq:probability:conjugate_priors:mult_likelihood:posterior} below.

@@

In general, there are two main views to probability and statistics. These include the \textit{frequentist} and the \textit{Bayesian} view of statistics. Naturally, Bayesian statistics is based on \index{Bayes' theorem}Bayes' theorem as was presented in Section \ref{sec:probability:bayes_theorem}. Bayesian statistics describe the probability of an event in terms of some belief, based on previous knowledge of the event, and the conditions under which the event happens~\cite{ref:hackenberger:2019}. To introduce the concept of Bayesian inference and analysis, one first need to understand the difference between these two approaches. Bayesian statistics out-date the frequentist approach, but lacked interest in the early days, partly because of the limited applications where the conjugate priors where known~\cite{ref:hackenberger:2019}. More recent advancements in mathematical methods popularised the Bayesian approach again. A notable contribution to this switch, was the development of \acf{MCMC} in the 1950's. This family of algorithms allowed for the construction of random sampling algorithms from a probability distribution, which allows for the calculation of Bayesian hierarchical models~\cite{ref:hackenberger:2019}. Soon after followed one of the earliest papers that use Bayesian statistics in the field of medicine in 1982~\cite{ref:ashby:2006}.

@@

Finally, the posterior density of $\theta \vert y_{1}, y_{2}, \dots, y_{n}$, according to \index{Bayes' theorem}Bayes' theorem, is given in Equation~\eqref{eq:probability:bayesian_statistic:bayesian_analysis:posterior_density} below.
